---
title: "SIG06 DEseq Analysis"
author: "Eric Y. Wang"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    html_preview: false
  html_notebook:
    toc: true
    toc_float: true
---

```{r setup}
library(tidyverse)
library(ggplot2)
library(cowplot)
library(patchwork)
library(DESeq2)
library(ComplexHeatmap)
library(IHW)
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
source("../functions/plotting_fxns.R")
theme_set(theme_Publication())
```

### [Import Data]{.underline}

```{r}
data <- read.csv("processing_outs/count_matrix_umiDeDup_SIG06.csv", row.names = "X")

metadata <- read_csv("processing_outs/processed_metadata_SIG06.csv")

# import feature names
featureNames <- read_csv("processing_outs/featureNames_SIG06.csv")
featureNames <- select(featureNames, -category)
```

### [Normalization and Diagnostic Plots]{.underline}

```{r}
# prepare metadata for DESeq
metaDF <- as.data.frame(metadata[,-1])
rownames(metaDF) <- metadata$sample_ID

# match matrix and metadata
data <- data[,rownames(metaDF)]
```

```{r}
# create DEseq object
dds <- DESeqDataSetFromMatrix(data,
                              colData = metaDF,
                              design = ~ treatment + replicate)

# prefilter dds
# remove any gene with fewer than 5 counts in the smallest group size
smallestGroupSize <- 3
keep <- rowSums(counts(dds) >= 2) >= smallestGroupSize
dds <- dds[keep,]
```

#### Export normalized counts

```{r}
dds <- estimateSizeFactors(dds)
```

```{r}
normCounts <- counts(dds, normalized=T) %>%
  as_tibble(rownames = "ensembl_ID") %>%
  left_join(featureNames)

write_csv(normCounts, "analysis_outs/counts_norm_DEseq2.csv")
```

#### Assess sample similarity

First, we'll perform PCA analysis to look at the overall similarity between samples.

```{r}
# perform VST normalization
# essentially normalizes to library size while stabilizing variance for lowly expressed genes
ddsNorm <- vst(dds)
```

```{r, fig.height=15, fig.width=18}
p1 <- DESeq2::plotPCA(ddsNorm, intgroup = "plate", ntop=500) + theme(aspect.ratio = 1) +
  scale_color_brewer(palette = "Dark2") +
  ggtitle("PC1 PC2 by plate")
p2 <- DESeq2::plotPCA(ddsNorm, intgroup = "plate", ntop=500, pcsToUse = 3:4) + theme(aspect.ratio = 1) +
  scale_color_brewer(palette = "Dark2") +
  ggtitle("PC3 PC4 by plate")
p3 <- DESeq2::plotPCA(ddsNorm, intgroup = "plate", ntop=500, pcsToUse = 5:6) + theme(aspect.ratio = 1) +
  scale_color_brewer(palette = "Dark2") +
  ggtitle("PC5 PC6 by plate")
p4 <- DESeq2::plotPCA(ddsNorm, intgroup = "plate", ntop=500, pcsToUse = 7:8) + theme(aspect.ratio = 1) +
  scale_color_brewer(palette = "Dark2") +
  ggtitle("PC7 PC8 by plate")

(p1+p2)/(p3+p4)
```

```{r, fig.height=15, fig.width=18}
p1 <- DESeq2::plotPCA(ddsNorm, intgroup = "replicate", ntop=500) + theme(aspect.ratio = 1) +
  scale_color_brewer(palette = "Dark2") +
  ggtitle("PC1 PC2 by replicate")
p2 <- DESeq2::plotPCA(ddsNorm, intgroup = "replicate", ntop=500, pcsToUse = 3:4) + theme(aspect.ratio = 1) +
  scale_color_brewer(palette = "Dark2") +
  ggtitle("PC3 PC4 by replicate")
p3 <- DESeq2::plotPCA(ddsNorm, intgroup = "replicate", ntop=500, pcsToUse = 5:6) + theme(aspect.ratio = 1) +
  scale_color_brewer(palette = "Dark2") +
  ggtitle("PC5 PC6 by replicate")
p4 <- DESeq2::plotPCA(ddsNorm, intgroup = "replicate", ntop=500, pcsToUse = 7:8) + theme(aspect.ratio = 1) +
  scale_color_brewer(palette = "Dark2") +
  ggtitle("PC7 PC8 by replicate")

(p1+p2)/(p3+p4)
```

```{r, fig.height=15, fig.width=18}
p1 <- DESeq2::plotPCA(ddsNorm, intgroup = "nCount_RNA", ntop=500) +
  theme(aspect.ratio = 1) +
  scale_color_viridis_c() +
  ggtitle("PC1 PC2 by nCount_RNA")
p2 <- DESeq2::plotPCA(ddsNorm, intgroup = "nCount_RNA", ntop=500, pcsToUse = 3:4) +
  theme(aspect.ratio = 1) +
  scale_color_viridis_c() +
  ggtitle("PC3 PC4 by nCount_RNA")
p3 <- DESeq2::plotPCA(ddsNorm, intgroup = "nCount_RNA", ntop=500, pcsToUse = 5:6) +
  theme(aspect.ratio = 1) +
  scale_color_viridis_c() +
  ggtitle("PC5 PC6 by nCount_RNA")
p4 <- DESeq2::plotPCA(ddsNorm, intgroup = "nCount_RNA", ntop=500, pcsToUse = 7:8) +
  theme(aspect.ratio = 1) +
  scale_color_viridis_c() +
  ggtitle("PC7 PC8 by nCount_RNA")

(p1+p2)/(p3+p4)
```

There don't appear to be any major outliers by plate, sequencing depth, or biological group. Plate 4 is a bit of an outlier since it has samples that aren't on plate 1/2/3 so I would expect it be a bit different than the others.

### [DEG Analysis]{.underline}

#### Create DESeq object

```{r}
# # perform DE analysis
# dds <- DESeq(dds, parallel = TRUE)
```

```{r}
# # save dds object
# saveRDS(dds, "analysis_outs/dds_object_SIG07.rds")

# import dds object
dds <- readRDS("analysis_outs/dds_object_SIG07.rds")
```


```{r}
# plot dispersion model to examine best fit
plotDispEsts(dds)
```

#### Calculate DEGs

Made a quick helper function to calculate DEGs.
```{r}
library(IHW)

# import feature names (used in function below)
featureNames <- read_csv("processing_outs/featureNames_SIG06.csv")
featureNames <- select(featureNames, -category)

# sample input can be single character or vector of characters
# control input must be single character, defaults to linker
iterative_DEG <- function(samples, control = "linker"){
  # empty tibble for full DEG list
  resAll <- tibble()
  
  for(ligand in samples){
    # perform IHW DEG analysis
    # don't perform independent filtering because using IHW
    res <- results(dds, contrast = c("treatment",ligand,control),
                 independentFiltering=F,
                 filterFun = ihw,
                 parallel = T)
    
    # perform LFC shrinkage and make tidy
    # pass res to lfcShrink
    resTidy <- lfcShrink(dds, type = "ashr", res=res) %>%
      as_tibble(rownames = "ensembl_ID") %>%
      left_join(featureNames) %>%
      mutate(treatment = ligand) %>%
      arrange(padj)
    
    # add to full DEG list
    resAll <- bind_rows(resAll, resTidy)
    
    return(resAll)
  }
}
```

```{r}
# select viral treatments groups
viralGroups <- dds$treatment %>% unique()
viralGroups <- viralGroups[grep("recomb|none|linker", viralGroups, invert = T)]
viralGroups
```

```{r}
res <- iterative_DEG("CXCL9")
res2 <- iterative_DEG("CXCL10")
res3 <- iterative_DEG("CXCL13")
res4 <- iterative_DEG("IL12recomb","none")
res5 <- iterative_DEG("IL6","linker")
res6 <- iterative_DEG("linker","none")
```





#### Variable Genes

```{r, fig.height=11, fig.width=9}
library(RColorBrewer)

# find top variable genes
topVarGenes <- head(order(rowVars(assay(ddsNorm)), decreasing = TRUE), 80)

# extract VST count matrix
mat  <- assay(ddsNorm)[topVarGenes, ]
# z-score VST counts
mat  <- t(scale(t(mat), center = T, scale = F))
anno <- as.data.frame(colData(ddsNorm)[, c("treatment","replicate")])

# replace rownames with gene names
rownames <- tibble(ensembl_ID = rownames(mat)) %>%
  left_join(featureNames)
rownames(mat) <- rownames$gene

pheatmap(mat, annotation_col = anno,
          annotation_colors = list(treatment = setNames(brewer.pal(10, "Paired"),levels(anno$treatment)),
                                   replicate = setNames(brewer.pal(3,"Dark2"),unique(anno$replicate))),
         main = "Centered VST normalized counts")
```

There is biological variability between different mice. For example, Il2 is consistently not upregulated in one mouse across all treatments.

```{r, fig.height=12, fig.width=8}
# extract VST count matrix
mat  <- assay(ddsNorm)[topVarGenes, ]
# z-score VST counts
mat  <- t(scale(t(mat), center = T, scale = F))
anno <- as.data.frame(colData(ddsNorm)[, c("treatment","replicate")])
# remove p123 titration samples
mat <- mat[,grep("^p123_1e",colnames(mat),invert = T)]
anno <- anno[grep("^p123_1e",rownames(anno),invert = T),]

# replace rownames with gene names
rownames <- tibble(ensembl_ID = rownames(mat)) %>%
  left_join(featureNames)
rownames(mat) <- rownames$gene

pheatmap(mat, annotation_col = anno,
          annotation_colors = list(treatment = setNames(brewer.pal(10, "Paired"),levels(anno$treatment)),
                                   replicate = setNames(brewer.pal(3,"Dark2"),unique(anno$replicate))),
         main = "Centered VST normalized counts")
```

```{r, fig.height=8, fig.width=6}
# find top variable genes
topVarGenes <- head(order(rowVars(assay(ddsNorm)), decreasing = TRUE), 30)

# extract VST count matrix
mat  <- assay(ddsNorm)[topVarGenes, ]
# z-score VST counts
mat  <- t(scale(t(mat), center = T, scale = F))
anno <- as.data.frame(colData(ddsNorm)[, c("treatment","replicate")])
# remove p123 titration samples
mat <- mat[,grep("^p123_1e",colnames(mat),invert = T)]
anno <- anno[grep("^p123_1e",rownames(anno),invert = T),]

# replace rownames with gene names
rownames <- tibble(ensembl_ID = rownames(mat)) %>%
  left_join(featureNames)
rownames(mat) <- rownames$gene

pheatmap(mat, annotation_col = anno,
          annotation_colors = list(treatment = setNames(brewer.pal(10, "Paired"),levels(anno$treatment)),
                                   replicate = setNames(brewer.pal(3,"Dark2"),unique(anno$replicate))),
         main = "Centered VST normalized counts")
```















