---
title: "UMAP and Clustering"
author: "Ian Zumpano"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    html_preview: false
  html_notebook:
    toc: true
    toc_float: true
---

```{r setup}
library(tidyverse)
library(ggplot2)
library(cowplot)
library(patchwork)
library(DESeq2)
library(ComplexHeatmap)
library(IHW)
library(RColorBrewer)
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
source("../functions/plotting_fxns.R")
theme_set(theme_Publication())
```

### [Import Data]{.underline}

```{r}
data <- read.csv("../processing_outs/count_matrix_umiDeDup_SIG06.csv", row.names = "X")

metadata <- read_csv("../processing_outs/processed_metadata_SIG06.csv")

# import feature names
featureNames <- read_csv("../processing_outs/featureNames_SIG06.csv")
featureNames <- select(featureNames, -category)
```

### [Normalization and Diagnostic Plots]{.underline}

```{r}
# prepare metadata for DESeq
metaDF <- as.data.frame(metadata[,-1])
rownames(metaDF) <- metadata$sample_ID

# match matrix and metadata
data <- data[,rownames(metaDF)]
```

#### Read normalized counts

```{r}
normCounts <- read.csv("../analysis_outs/counts_norm_DEseq2.csv")
```

#### Create DESeq object

```{r}
# import dds object
dds <- readRDS("/Users/izumpano/Desktop/Eric Analyses/SIG06/analysis_outs/dds_object_SIG06.rds")
```

#### Create Treatment Groups 

```{r}
# select viral treatments groups
viralGroups <- dds$treatment %>% unique()
viralGroups <- viralGroups[grep("recomb|none|linker", viralGroups, invert = T)]
viralGroups
```

```{r}
# select recombinant treatments groups
recombGroups <- dds$treatment %>% unique()
recombGroups <- grep("recomb", recombGroups, value = T)
recombGroups
```

```{r}
# select linkers
linkerGroups <- dds$treatment %>% unique()
linkerGroups <- grep("linker", linkerGroups, value = T)
linkerGroups
```

### Load DEG Analysis

```{r}
res_viral <- read.csv("../analysis_outs/res_viral.csv")
```

### UMAP Analysis

#### Perform UMAP

```{r}
library(umap)

dds_norm <- vst(dds)

# First we are going to retrieve the normalized data
# from the `DESeqDataSet` object using the `assay()` function
normalized_counts_umap <- assay(dds_norm) %>%
  t() # We need to transpose this data so each row is a sample

# Now perform UMAP on the normalized data
umap_results <- umap::umap(normalized_counts_umap)
```

#### Prepare data frame for plotting

```{r}
# Make into data frame for plotting with `ggplot2`
# The UMAP values we need for plotting are stored in the `layout` element
umap_plot_df <- data.frame(umap_results$layout) %>%
  # Turn sample IDs stored as row names into a column
  tibble::rownames_to_column("sample_ID") %>%
  # Add the metadata into this data frame; match by sample IDs
  dplyr::inner_join(metadata, by = "sample_ID")


#Create Well to row col dataset
umap_plot_df_well_rowcol <- umap_plot_df %>%
  mutate(
    well_row = sub("\\d+", "", well),       # Extract the letter (row)
    well_column = as.numeric(sub("\\D+", "", well)) # Extract the number (column)
  )

```

#### Create UMAP Plots

```{r, fig.height=10, fig.width=12}
library(ggrepel)
library(patchwork)
library(ggrepel)

p1 <- ggplot(umap_plot_df, aes(x = X1, y = X2, color = replicate)) +
  geom_point(shape = 15) + 
  ggrepel::geom_text_repel(data = umap_plot_df) +
  labs(title="UMAP by Group")

p2 <- ggplot(umap_plot_df, aes(x = X1, y = X2, color = plate)) +
  geom_point(shape = 15) + 
  labs(title="UMAP by RNA-seq Plate")

p3 <- ggplot(umap_plot_df_well_rowcol, aes(x = X1, y = X2, color = well_row)) +
  geom_point(shape = 15) + 
  labs(title="UMAP by Row")

p4 <- ggplot(umap_plot_df_well_rowcol, aes(x = X1, y = X2, color = factor(well_column))) +
  geom_point(shape = 15) + 
  labs(title="UMAP by Column")

p5 <- ggplot(umap_plot_df_well_rowcol, aes(x = X1, y = X2, color = factor(treatment))) +
  geom_point(shape = 15) + 
  labs(title="UMAP by Treatment")



p1+p2+p3+p4+plot_layout(ncol=2)
# highlight which cells are the linker only cells
# run a clustering algorithm on data to identify discrete clusters, then project clusters onto the umap to see which ligands are enriched clusters
# simple heirarchical clustering
```

```{r, fig.height= 12, fig.width= 18}
p5
```


### Run Clustering Analysis

```{r, fig.width=24, fig.height=30}
library(ggdendro)

hc <-  normalized_counts_umap %>%
   dist(method = "euclidean") %>%
   hclust()

ggdendrogram(hc, rotate=TRUE)

#add in metadata
hc_meta <- hc
hc_meta$labels <- str_c(metadata$treatment, ":", metadata$nCount_RNA)
ggdendrogram(hc_meta, rotate=TRUE)

```

#### Questions from this plot:

why is CXCL13_group2 basal from the rest of the plot?

### Overlay Clustering on UMAP

```{r}
#look for ligands that have one/no differentially expressed genes

dark2_pal <- brewer.pal(8, "Dark2")

# Define the number of clusters
num_clusters <- 5
clusters <- cutree(hc, k = num_clusters)
clusters_meta <- cutree(hc_meta, k = num_clusters)

umap_cluster <- umap_plot_df
umap_cluster$cluster <- as.factor(clusters)  # Add the cluster assignments to UMAP results

umap_cluster$cluster_meta <- as.factor(clusters_meta)

# Plot UMAP with clusters
ggplot(umap_cluster, aes(x = X1, y = X2, color = cluster)) +
  geom_point(size = 2, shape = 15) +
  labs(title = "UMAP Projection with Hierarchical Clusters", x = "UMAP 1", y = "UMAP 2") +
  theme_minimal() +
  scale_color_manual(values = c("salmon", "navyblue", "seagreen", "darkgoldenrod1", "darkorchid")) 

# Plot UMAP with clusters
ggplot(umap_cluster, aes(x = X1, y = X2, color = cluster_meta)) +
  geom_point(size = 2, shape = 15) +
  labs(title = "UMAP Projection with Hierarchical Clusters (Metadata)", x = "UMAP 1", y = "UMAP 2") +
  theme_minimal() +
  scale_color_manual(values = c("salmon", "navyblue", "seagreen", "darkgoldenrod1", "darkorchid")) 


# try overlaying ligands onto umap
# take a closer look at PCs, some ligands that have strong signalling effect, look at pcas only on genes that dont have strong DEGs (null genes subset), might reveal what is driving batch effect more
# minimal distortion embedding
# focus on 2 and 4, try to figure out what covars are driving variation
# want to see how mych overlap genes that are DE between viral and DE for recomb

# for 2, make seperate notebook
```

```{r}
# Non DE Genes

dds_null <- dds_norm[null_genes, ]

#take the ligands that dont have differentially expressed genes and use those ligands to find principal components

non_DEG_treatment <- res_viral %>% filter(padj < 0.1) %>% group_by(treatment) %>% summarize(DEG_count = n() < 10) %>% arrange(desc(DEG_count)) %>% filter(DEG_count == TRUE) %>% pull(treatment)

filtered_coldata <- colData(dds_norm) %>%
  as.data.frame() %>%
  filter(treatment %in% non_DEG_treatment | treatment == "linker" )

dds_null <- dds_norm[, rownames(filtered_coldata)]

colData(dds_null)$shape_group <- ifelse(colData(dds_null)$treatment == "linker", "linker", "non_linker")

# Just use the linkers and try pca on those
```


### linear regression

```{r}
# Extract PCA data from DESeq2 plotPCA
pca_data1 <- DESeq2::plotPCA(dds_null, intgroup = c("treatment", "plate", "replicate"), pcsToUse = 1:2, ntop = 500, returnData = TRUE)
pca_data2 <- DESeq2::plotPCA(dds_null, intgroup = c("treatment", "plate", "replicate"), pcsToUse = 3:4, ntop = 500, returnData = TRUE)
pca_data3 <- DESeq2::plotPCA(dds_null, intgroup = c("treatment", "plate", "replicate"), pcsToUse = 5:6, ntop = 500, returnData = TRUE)
pca_data4 <- DESeq2::plotPCA(dds_null, intgroup = c("treatment", "plate", "replicate"), pcsToUse = 7:8, ntop = 500, returnData = TRUE)


fit1 <- lm(PC1 ~ plate + replicate, data = pca_data1)
summary(fit1)

fit2 <- lm(PC2 ~ plate + replicate, data = pca_data1)
summary(fit2)

fit3 <- lm(PC3 ~ plate + replicate, data = pca_data2)
summary(fit3)

fit4 <- lm(PC4 ~ plate + replicate, data = pca_data2)
summary(fit4)

fit5 <- lm(PC5 ~ plate + replicate, data = pca_data3)
summary(fit5)

fit6 <- lm(PC6 ~ plate + replicate, data = pca_data3)
summary(fit6)

fit7 <- lm(PC7 ~ plate + replicate, data = pca_data4)
summary(fit7)

fit8 <- lm(PC8 ~ plate + replicate, data = pca_data4)
summary(fit8)
```


### Plot PCs of Null Genes

#### Plate

```{r, fig.height=15, fig.width=18}
plot_plate_pca <- function(pcs){
pca_data <- DESeq2::plotPCA(
  dds_null, 
  intgroup = c("plate", "shape_group"), 
  ntop = 500, 
  pcsToUse = pcs,
  returnData = TRUE
)

# Extract percent variance for PC1 and PC2
percentVar <- round(100 * attr(pca_data, "percentVar"), 1)

# Plot using ggplot2
library(ggplot2)
ggplot(pca_data,  aes(x = !!sym(paste0("PC", pcs[1])), y = !!sym(paste0("PC", pcs[2])))) +
  geom_point(
    aes(color = plate, shape = shape_group),
    size = 3
  ) +
  scale_color_brewer(palette = "Dark2") + # Use the specified color palette
  scale_shape_manual(values = c("linker" = 17, "non_linker" = 16)) + # Shapes for linker/non-linker
  theme(aspect.ratio = 1) +
  labs(
    title = "PCA Plot by Plate and Linker Status",
    x = paste0("PC", pcs[1], ": ", percentVar[1], "% variance"),
    y = paste0("PC", pcs[2], ": ", percentVar[2], "% variance"),
    color = "Plate",
    shape = "Linker Status"
  )
}

p1 <- plot_plate_pca(pcs = 1:2)
p2 <- plot_plate_pca(pcs = 3:4)
p3 <- plot_plate_pca(pcs = 5:6)
p4 <- plot_plate_pca(pcs = 7:8)


# Arrange plots into a grid
(p1 + p2) / (p3 + p4)

```

looks like some grouping by plate across pc2 and pc3

#### Replicate

```{r, fig.height=15, fig.width=18}
plot_rep_pca <- function(pcs){
pca_data <- DESeq2::plotPCA(
  dds_null, 
  intgroup = c("replicate", "shape_group"), 
  ntop = 500, 
  pcsToUse = pcs,
  returnData = TRUE
)

# Extract percent variance for PC1 and PC2
percentVar <- round(100 * attr(pca_data, "percentVar"), 1)

# Plot using ggplot2
library(ggplot2)
ggplot(pca_data,  aes(x = !!sym(paste0("PC", pcs[1])), y = !!sym(paste0("PC", pcs[2])))) +
  geom_point(
    aes(color = replicate, shape = shape_group),
    size = 3
  ) +
  scale_color_brewer(palette = "Dark2") + # Use the specified color palette
  scale_shape_manual(values = c("linker" = 17, "non_linker" = 16)) + # Shapes for linker/non-linker
  theme(aspect.ratio = 1) +
  labs(
    title = "PCA Plot by Replicate and Linker Status",
    x = paste0("PC", pcs[1], ": ", percentVar[1], "% variance"),
    y = paste0("PC", pcs[2], ": ", percentVar[2], "% variance"),
    color = "Replicate",
    shape = "Linker Status"
  )
}

p1 <- plot_rep_pca(pcs = 1:2)
p2 <- plot_rep_pca(pcs = 3:4)
p3 <- plot_rep_pca(pcs = 5:6)
p4 <- plot_rep_pca(pcs = 7:8)


# Arrange plots into a grid
(p1 + p2) / (p3 + p4)
# look at pc loadings, see which genes are dribing most of the pc
```

#### nCount_RNA

```{r, fig.height=15, fig.width=18}
plot_count_pca <- function(pcs){
pca_data <- DESeq2::plotPCA(
  dds_null, 
  intgroup = c("nCount_RNA", "shape_group"), 
  ntop = 500, 
  pcsToUse = pcs,
  returnData = TRUE
)

pca_data <- pca_data %>%
    mutate(log10_nCount_RNA = log10(nCount_RNA))

# Extract percent variance for PC1 and PC2
percentVar <- round(100 * attr(pca_data, "percentVar"), 1)

# Plot using ggplot2
library(ggplot2)
ggplot(pca_data,  aes(x = !!sym(paste0("PC", pcs[1])), y = !!sym(paste0("PC", pcs[2])))) +
  geom_point(
    aes(color = nCount_RNA, shape = shape_group),
    size = 3
  ) +
  scale_color_viridis_c() + # Use the specified color palette
  scale_shape_manual(values = c("linker" = 17, "non_linker" = 16)) + # Shapes for linker/non-linker
  theme(aspect.ratio = 1) +
  labs(
    title = "PCA Plot by log10 nCount_RNA and Linker Status",
    x = paste0("PC", pcs[1], ": ", percentVar[1], "% variance"),
    y = paste0("PC", pcs[2], ": ", percentVar[2], "% variance"),
    color = "log10(nCount_RNA)",
    shape = "Linker Status"
  )
}

p1 <- plot_count_pca(pcs = 1:2)
p2 <- plot_count_pca(pcs = 3:4)
p3 <- plot_count_pca(pcs = 5:6)
p4 <- plot_count_pca(pcs = 7:8)


# Arrange plots into a grid
(p1 + p2) / (p3 + p4)

```

### PCA of full data by well row and column

#### Row

```{r, fig.height=15, fig.width=18}
colData(dds_norm)$shape_group <- ifelse(colData(dds_norm)$treatment == "linker", "linker", "non_linker")

plot_row_pca <- function(pcs) {
  # Add well_row to colData if not already present
  colData(dds_norm)$well_row <- gsub("[0-9]", "", colData(dds_norm)$well)

  # Extract PCA data with well_row as a grouping variable
  pca_data <- DESeq2::plotPCA(
    dds_norm, 
    intgroup = c("well_row", "shape_group"), 
    ntop = 500, 
    pcsToUse = pcs,
    returnData = TRUE
  )

  # Extract percent variance for the PCs
  percentVar <- round(100 * attr(pca_data, "percentVar"), 1)

  # Plot using ggplot2
  ggplot(pca_data, aes(x = !!sym(paste0("PC", pcs[1])), y = !!sym(paste0("PC", pcs[2])))) +
    geom_point(
      aes(color = well_row, shape = shape_group),
      size = 3
    ) +
    scale_color_brewer(palette = "Dark2") + # Use the specified color palette
    scale_shape_manual(values = c("linker" = 17, "non_linker" = 16)) + # Shapes for linker/non-linker
    theme(aspect.ratio = 1) +
    labs(
      title = "PCA Plot by Well Row and Linker Status",
      x = paste0("PC", pcs[1], ": ", percentVar[1], "% variance"),
      y = paste0("PC", pcs[2], ": ", percentVar[2], "% variance"),
      color = "Well Row",
      shape = "Linker Status"
    )
}

# Generate PCA plots for different PC combinations
p1 <- plot_row_pca(pcs = 1:2)
p2 <- plot_row_pca(pcs = 3:4)
p3 <- plot_row_pca(pcs = 5:6)
p4 <- plot_row_pca(pcs = 7:8)

# Arrange the plots into a grid
(p1 + p2) / (p3 + p4)
```


#### Column

```{r, fig.height=15, fig.width=18}
plot_col_pca <- function(pcs) {
  # Add well_column to colData if not already present
  colData(dds_norm)$well_column <- gsub("[A-Za-z]", "", colData(dds_norm)$well)

  # Extract PCA data with well_column as a grouping variable
  pca_data <- DESeq2::plotPCA(
    dds_norm, 
    intgroup = c("well_column", "shape_group"), 
    ntop = 500, 
    pcsToUse = pcs,
    returnData = TRUE
  )

  # Extract percent variance for the PCs
  percentVar <- round(100 * attr(pca_data, "percentVar"), 1)

  # Plot using ggplot2
  ggplot(pca_data, aes(x = !!sym(paste0("PC", pcs[1])), y = !!sym(paste0("PC", pcs[2])))) +
    geom_point(
      aes(color = well_column, shape = shape_group),
      size = 3
    ) +
    scale_color_brewer(palette = "Set3") + # Use a viridis color palette for numeric well columns
    scale_shape_manual(values = c("linker" = 17, "non_linker" = 16)) + # Shapes for linker/non-linker
    theme(aspect.ratio = 1) +
    labs(
      title = "PCA Plot by Well Column and Linker Status",
      x = paste0("PC", pcs[1], ": ", percentVar[1], "% variance"),
      y = paste0("PC", pcs[2], ": ", percentVar[2], "% variance"),
      color = "Well Column",
      shape = "Linker Status"
    )
}

# Generate PCA plots for different PC combinations
p1 <- plot_col_pca(pcs = 1:2)
p2 <- plot_col_pca(pcs = 3:4)
p3 <- plot_col_pca(pcs = 5:6)
p4 <- plot_col_pca(pcs = 7:8)

# Arrange the plots into a grid
(p1 + p2) / (p3 + p4)
```

# what genes make up the pcs that were correlated with plate and replicate variation

### Genes Driving PCs

```{r}
# Extract normalized counts or variance-stabilized data
vsd_data <- assay(dds_null)  # Use DESeq2's variance-stabilizing transformation

# Perform PCA
pca_res <- prcomp(t(vsd_data))  # Transpose so samples are rows and genes are columns
```

#### Extract loadings

```{r, fig.height = 8, fig.width = 10}
get_pc_genes <- function(pc) {
  # Extract loadings
  loadings <- pca_res$rotation  # Genes x PCs
  
  # Get top 10 positive and top 10 negative loadings
  pc_top_genes_pos <- loadings[, pc] %>%
    sort(decreasing = TRUE) %>%
    head(20)
  
  pc_top_genes_neg <- loadings[, pc] %>%
    sort(decreasing = FALSE) %>%
    head(20)
  
  # Combine positive and negative loadings
  pc_top_genes <- c(pc_top_genes_pos, pc_top_genes_neg)
  
  # Create a data frame for the top genes
  top_genes_df <- data.frame(
    ensembl_ID = names(pc_top_genes),
    Loading = pc_top_genes
  )
  
  # Map Ensembl IDs to gene names using featureNames
  top_genes_named <- top_genes_df %>%
    left_join(featureNames, by = "ensembl_ID")
  
  # Add a column for the loading type (Positive/Negative)
  top_genes_named <- top_genes_named %>%
    mutate(LoadingType = ifelse(Loading > 0, "Positive", "Negative"))
  
  # Bar plot with gene names, color-coded by loading type
  ggplot(top_genes_named, aes(x = reorder(gene, Loading), y = Loading, fill = LoadingType)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    scale_fill_manual(values = c("Positive" = "steelblue", "Negative" = "darkred")) +
    labs(
      title = paste0("Top 20 Positive and Negative Genes Driving ", pc),
      x = "Gene",
      y = "Loading",
      fill = "Loading Type"
    ) +
    theme_minimal()
}

# Generate plots for multiple PCs
get_pc_genes(pc = "PC1")
get_pc_genes(pc = "PC2")
get_pc_genes(pc = "PC3")
get_pc_genes(pc = "PC4")
get_pc_genes(pc = "PC5")
get_pc_genes(pc = "PC6")
get_pc_genes(pc = "PC7")
get_pc_genes(pc = "PC8")
```

