---
title: "Correlation Heatmaps"
author: "Ian Zumpano"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    html_preview: false
  html_notebook:
    toc: true
    toc_float: true
---

```{r setup}
library(tidyverse)
library(ggplot2)
library(cowplot)
library(patchwork)
library(DESeq2)
library(ComplexHeatmap)
library(IHW)
library(RColorBrewer)
library(viridis)
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
source("../functions/plotting_fxns.R")
theme_set(theme_Publication())
```

### [Import Data]{.underline}

```{r}
data <- read.csv("../processing_outs/count_matrix_umiDeDup_SIG06.csv", row.names = "X")

metadata <- read_csv("../processing_outs/processed_metadata_SIG06.csv")

# import feature names
featureNames <- read_csv("../processing_outs/featureNames_SIG06.csv")
featureNames <- select(featureNames, -category)
```

### [Normalization and Diagnostic Plots]{.underline}

```{r}
# prepare metadata for DESeq
metaDF <- as.data.frame(metadata[,-1])
rownames(metaDF) <- metadata$sample_ID

# match matrix and metadata
data <- data[,rownames(metaDF)]
```

#### Read normalized counts

```{r}
normCounts <- read.csv("../analysis_outs/counts_norm_DEseq2.csv")
```

#### Create DESeq object

```{r}
# import dds object
dds <- readRDS("/Users/izumpano/Desktop/Eric Analyses/SIG06/analysis_outs/dds_object_SIG06.rds")
```

#### Create Treatment Groups 

```{r}
# select viral treatments groups
viralGroups <- dds$treatment %>% unique()
viralGroups <- viralGroups[grep("recomb|none|linker", viralGroups, invert = T)]
viralGroups
```

```{r}
# select recombinant treatments groups
recombGroups <- dds$treatment %>% unique()
recombGroups <- grep("recomb", recombGroups, value = T)
recombGroups
```

```{r}
# select linkers
linkerGroups <- dds$treatment %>% unique()
linkerGroups <- grep("linker", linkerGroups, value = T)
linkerGroups
```

### Load DEG Analysis

```{r}
res_viral <- read.csv("../analysis_outs/res_viral.csv")
```

### Plot Correlation Heatmaps

### Prepare annotation and metadata

```{r}
metadata$treatment_type <- ifelse(metadata$treatment == "linker", "Linker",
                                    ifelse(metadata$treatment == "none", "None", "Other")) %>%  as_factor()

metadata$log_nCount_RNA <- log10(metadata$nCount_RNA)

annotation <- metadata %>% 
	select(sample_ID, replicate, plate, treatment_type, log_nCount_RNA) %>% data.frame(row.names = "sample_ID")
  
normCounts_rows <-normCounts %>% column_to_rownames("ensembl_ID") %>% select(-gene)

  
```

#### Create Helper Function to Plot Heatmaps

```{r}
# Define annotation colors
annotation_colors <- list(
  replicate = c(group1 = "darkblue", group2 = "darkgreen", group3 = "darkred"),
  plate = c(plate1 = "purple", plate2 = "orange", plate3 = "yellow", plate4 = "black"),
  treatment_type = c(Linker = "blue", None = "red", Other = "gray")
)

plot_ligand_correlation_heatmap <- function(norm_counts, metadata, annotation, method = "pearson", color_palette = "RdYlBu", title = "") {
  # Compute correlation matrix
  correlation_matrix <- cor(norm_counts, method = method)
  
  # Plot heatmap
  pheatmap(
    correlation_matrix,
    annotation_col = annotation,
    annotation_row = annotation,
    annotation_colors = annotation_colors,
    show_rownames = FALSE,
    show_colnames = FALSE,
    color = viridis(100, option = "magma"),
    cluster_rows = TRUE,
    cluster_cols = TRUE,
    legend = TRUE,
    main = paste(method, "Correlation Heatmap of Ligands", title)
  )
}
```

#### Heatmap for all ligands using vst normalized counts

```{r, fig.width=18, fig.height=12}
dds_norm <- vst(dds)
normCounts_vst <- assay(dds_norm)

normCounts_vst <- as.data.frame(normCounts_vst)

plot_ligand_correlation_heatmap(normCounts_vst, metadata, annotation = annotation, title = "(VST Normalized Counts)")
```

### Heatmap of Ligands by Number of DEGs

#### 10 DEGs

```{r, fig.width=18, fig.height=12}
library(stringi)
library(stringr)

# more than 10 DEGs
DEG_10 <- res_viral %>% filter(padj < 0.1) %>% group_by(treatment) %>% summarize(DEG_count = n() >= 10) %>% filter(DEG_count == TRUE) %>% pull(treatment) %>% unique()

# Helper function to extract columns corresponding to DE ligands
filter_columns_by_ligands <- function(dataset, top_ligands) {
  # Extract the column names that match the criteria
  columns_to_keep <- names(dataset) %>%
    keep(~ str_split_fixed(.x, "_", 2)[1] %in% top_ligands)
  
  # Select only these columns
  dataset %>% select(all_of(columns_to_keep))
} 

# Filter the dataset
normCounts_DEG_10 <- filter_columns_by_ligands(normCounts_rows, DEG_10)

annotation_DEG_10 <- annotation[rownames(annotation) %in% colnames(normCounts_DEG_10), , drop = FALSE]

plot_ligand_correlation_heatmap(normCounts_DEG_10, metadata, annotation = annotation_DEG_10, title = "(Ligands with 10+ DEGs)")
```

#### 15+ DEGs

```{r, fig.width=18, fig.height=12}
# Filter the dataset
DEG_15 <- res_viral %>% filter(padj < 0.1) %>% group_by(treatment) %>% summarize(DEG_count = n() >= 15) %>% filter(DEG_count == TRUE) %>% pull(treatment) %>% unique()

normCounts_DEG_15 <- filter_columns_by_ligands(normCounts_rows, DEG_15)

annotation_DEG_15 <- annotation[rownames(annotation) %in% colnames(normCounts_DEG_15), , drop = FALSE]

plot_ligand_correlation_heatmap(normCounts_DEG_15, metadata, annotation = annotation_DEG_15, title = "(Ligands with 15+ DEGs)")
```

#### 50+ DEGs

```{r, fig.width=18, fig.height=12}
# Filter the dataset
DEG_50 <- res_viral %>% filter(padj < 0.1) %>% group_by(treatment) %>% summarize(DEG_count = n() >= 50) %>% filter(DEG_count == TRUE) %>% pull(treatment) %>% unique()

normCounts_DEG_50 <- filter_columns_by_ligands(normCounts_rows, DEG_50)

annotation_DEG_50 <- annotation[rownames(annotation) %in% colnames(normCounts_DEG_50), , drop = FALSE]

plot_ligand_correlation_heatmap(normCounts_DEG_50, metadata, annotation = annotation_DEG_50, title = "(Ligands with 50+ DEGs)")
```

#### 500+ DEGs

```{r, fig.width=18, fig.height=12}
# Filter the dataset
DEG_500 <- res_viral %>% filter(padj < 0.1) %>% group_by(treatment) %>% summarize(DEG_count = n() >= 500) %>% filter(DEG_count == TRUE) %>% pull(treatment) %>% unique()

normCounts_DEG_500 <- filter_columns_by_ligands(normCounts_rows, DEG_500)

annotation_DEG_500 <- annotation[rownames(annotation) %in% colnames(normCounts_DEG_500), , drop = FALSE]

plot_ligand_correlation_heatmap(normCounts_DEG_500, metadata, annotation = annotation_DEG_500, title = "(Ligands with 500+ DEGs)")

```

#### Correlation heatmap of most highly variable genes (not working yet)

```{r, fig.width=18, fig.height=12}
library(matrixStats)

# Compute row-wise variance
row_variances <- rowVars(as.matrix(normCounts_vst))

# Find the threshold for the top 10% most variable genes
threshold <- quantile(row_variances, 0.90)

# Get the indices of rows that meet or exceed the threshold
top_variable_indices <- which(row_variances >= threshold)

# Extract the row names corresponding to the top 10% most variable genes
top_hvg <- rownames(normCounts_vst)[top_variable_indices]

top_hvg_ligands <- res_viral %>% filter(ensembl_ID %in% top_hvg) %>% pull(treatment) %>% unique()

normCounts_HVG <- filter_columns_by_ligands(normCounts_vst, top_hvg_ligands)


annotation_HVG <- annotation[rownames(annotation) %in% colnames(normCounts_HVG), , drop = FALSE]

plot_ligand_correlation_heatmap(normCounts_HVG, metadata, annotation = annotation_HVG, title = "(Top HVG Ligands)")

```

#### Average of Replicates Heatmap

```{r, fig.width=20, fig.height=14}
#make separate heatmap instead of plotting all replicates plot avg expression of each replicate

library(purrr)

#treatment_groups <- names(normCounts_vst) %>% str_extract("^[^_]+") %>% unique()

# Function to calculate row-wise means for each treatment group
library(dplyr)
library(stringr)

# Function to calculate row-wise means based on treatment groups
rowwise_means <- function(dataset, treatment_groups) {
  # Initialize an empty dataframe to store the results
  result <- data.frame(row.names = rownames(dataset))
  
  # Loop over each treatment in treatment_groups
  for (treatment in treatment_groups) {
    # Select columns matching the current treatment
    matching_cols <- dataset %>%
      select(starts_with(treatment))
    
    # Calculate row-wise means for these columns
    row_means <- rowMeans(matching_cols, na.rm = TRUE)
    
    # Add the row means as a new column in the result dataframe
    result[[treatment]] <- row_means
  }
  
  return(result)
}


normCounts_AVG <- rowwise_means(normCounts_vst, viralGroups)

#plot heatmap by hand bc no annotation
correlation_matrix <- cor(normCounts_AVG, method = "pearson")
  
  # Plot heatmap
pheatmap(
    correlation_matrix,
    show_rownames = TRUE,
    show_colnames = TRUE,
    color =  viridis(100, option = "magma"),
    cluster_rows = TRUE,
    cluster_cols = TRUE,
    legend = TRUE,
    main = "Pearson Correlation Heatmap of Ligands (Mean Across Replicates)"
  )

# remove recomb and none
# hvgs averaged across groups


#top_hvg_ligands

```

#### HVGs averaged across replicates

```{r}
normCounts_AVG %>% select(names(normCounts_AVG) %in% top_hvg_ligands)

setdiff(treatment_groups, top_hvg_ligands)
```
```{r}
head(res_viral)
```


