---
title: "Factor Analysis"
author: "Ian Zumpano"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_float: true
  github_document:
    toc: true
    html_preview: false
---

```{r setup}
library(tidyverse)
library(ggplot2)
library(ggrepel)
library(cowplot)
library(patchwork)
library(DESeq2)
library(ComplexHeatmap)
library(IHW)
library(RColorBrewer)
library(FactoMineR)   # For PCA and factor analysis
library(factoextra)   # For PCA visualization
library(psych)        # For factor analysis
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
source("functions/plotting_fxns.R")
#source("../functions/plotting_fxns.R")
theme_set(theme_Publication())
```

### [Import Data]{.underline}

```{r}
# import stuff on windows
data <- read.csv("processing_outs/count_matrix_umiDeDup_SIG06.csv", row.names = "X")
metadata <- read_csv("processing_outs/processed_metadata_SIG06.csv")
# import feature names
featureNames <- read_csv("processing_outs/featureNames_SIG06.csv")
featureNames <- select(featureNames, -category)
# import DEGs
res_viral_sig <- read.csv("analysis_outs/res_viral_sig.csv")

# import stuff on mac
# data <- read.csv("../processing_outs/count_matrix_umiDeDup_SIG06.csv", row.names = "X")
# metadata <- read_csv("../processing_outs/processed_metadata_SIG06.csv")
# # import feature names
# featureNames <- read_csv("../processing_outs/featureNames_SIG06.csv")
# featureNames <- select(featureNames, -category)
# res_viral <- read.csv("../analysis_outs/res_viral.csv")
```

### [Normalization and Diagnostic Plots]{.underline}

```{r}
# prepare metadata for DESeq
metaDF <- as.data.frame(metadata[,-1])
rownames(metaDF) <- metadata$sample_ID
# match matrix and metadata
data <- data[,rownames(metaDF)]
```

#### Read normalized counts

```{r}
normCounts <- read.csv("../analysis_outs/counts_norm_DEseq2.csv")
```

#### Create DESeq object and VST normalized counts

```{r}
# import dds object
dds <- readRDS("analysis_outs/dds_object_SIG06.rds")
vst <- vst(dds)
norm_counts <- assay(vst)
```

### Filter only highly variable genes

```{r}
library(matrixStats)
row_variance <- rowVars(norm_counts)
hvgs <- norm_counts[row_variance > quantile(row_variance, 0.75), ]
# get number of HVGs
dim(hvgs)
```

#### Scaled data

```{r}
scaled_data <- t(scale(t(hvgs)))
```

### PCA

#### Run & Visualize PCA

```{r, fig.height = 12, fig.width=20}
# # PCA
# pca <- prcomp(t(norm_counts)[, 1:8], scale. = TRUE)
# pca_hvg <- prcomp(hvgs[, 1:8], scale. = TRUE)
# 
# 
# # Visualize PCA
# library(ggplot2)
# pca_df <- as.data.frame(pca$x)
# pca_df$treatment <- metadata$treatment
# ggplot(pca_df, aes(x = PC1, y = PC2, color = treatment)) +
#   geom_point(size = 3) +
#   theme_minimal() +
#   labs(title = "PCA on Highly Variable Genes")
# 
# pca_df <- as.data.frame(pca_hvg$x)
# pca_df$treatment <- metadata$treatment
# ggplot(pca_df, aes(x = PC1, y = PC2, color = treatment)) +
#   geom_point(size = 3) +
#   theme_minimal() +
#   labs(title = "PCA on Highly Variable Genes")

```

### Gene set enrichment analysis

#### All genes 

```{r}
# pc1_loadings <- data.frame(
#   gene = rownames(pca$rotation),
#   value = pca$rotation[, "PC1"]
# ) %>%
#   arrange(desc(abs(value)))
# 
# geneList <- pc1_loadings %>%
#   dplyr::select(gene, value) %>%
#   arrange(desc(value)) %>%  # Sort in decreasing order
#   deframe()  # Convert to named vector
# 
# mapped_genes <- bitr(names(geneList), fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Mm.eg.db)
# geneList <- deframe(mapped_genes)
# 
# # GSEA
# library(clusterProfiler)
# library(org.Mm.eg.db) # Mouse annotations
# 
# gsea_results <- clusterProfiler::gseGO(
#   geneList = geneList,
#   OrgDb = org.Mm.eg.db,
#   ont = "BP",
#   keyType = "ENSEMBL",
#   nPerm = 1000,            # Number of permutations
#   minGSSize = 5,          # Minimum size of gene sets
#   maxGSSize = 1000,         # Maximum size of gene sets
#   pvalueCutoff = 0.1,     # Adjusted p-value cutoff
#   verbose = TRUE
# )
# 
# # Visualize GSEA results
# dotplot(gsea_results)
# 
# mapped_genes <- bitr(names(geneList), fromType = "ENSEMBL", toType = "ENTREZID", OrgDb = org.Mm.eg.db)
# head(mapped_genes)
# 
# # Check overlap
# overlap <- sum(mapped_genes$SYMBOL %in% names(geneList))
# print(overlap)  # Number of overlapping genes

# 
# geneList <- geneList[names(geneList) %in% mapped_genes$ENSEMBL]
# names(geneList) <- mapped_genes$SYMBOL[match(names(geneList), mapped_genes$ENSEMBL)]
# 
# overlap <- sum(names(geneList) %in% mapped_genes$SYMBOL)
# print(overlap) 
# 
# mapped_genes <- bitr(
# names(geneList),
# fromType = "ENSEMBL",
# toType = "SYMBOL",
# OrgDb = org.Mm.eg.db
# )
```


```{r}
# dge_mapped_df <- data.frame(
#   gene_symbol = mapIds(
#     # Replace with annotation package for the organism relevant to your data
#     org.Mm.eg.db,
#     keys = res_viral$ensembl_ID,
#     # Replace with the type of gene identifiers in your data
#     keytype = "ENSEMBL",
#     # Replace with the type of gene identifiers you would like to map to
#     column = "SYMBOL",
#     # This will keep only the first mapped value for each Ensembl ID
#     multiVals = "first"))
# dge_mapped_d
#   # )  %>%
#   # # If an Ensembl gene identifier doesn't map to a gene symbol, drop that
#   # # from the data frame
#   # dplyr::filter(!is.na(gene_symbol)) %>%
#   # # Make an `Ensembl` column to store the rownames
#   # tibble::rownames_to_column("Ensembl") %>%
#   # # Now let's join the rest of the expression data
#   # dplyr::inner_join(res_viral, by = c("Ensembl" = "Gene")))
```



### Sparse PCA

<<<<<<< Updated upstream
#### 50 genes per component

```{r, fig.height = 12, fig.width=20}
library(mixOmics)
library(GGally)
library(ggrepel)

if (ncol(hvgs) == 330) {
  hvgs <- t(hvgs)
}

#Only want to do PCA on HVGs

# Run sparse PCA
spca_res <- spca(hvgs, ncomp = 16, keepX = c(50, 50, 50, 50, 50, 50, 50, 50, 50, 50, 50, 50, 50, 50, 50, 50))  # Select 50 genes per component


#plotIndiv(spca_res, group = metadata$treatment, legend = TRUE)

spca_scores_30 <- as.data.frame(spca_res$variates$X)   # Samples x Components

# Add metadata to sPCA scores
spca_scores_30$group <- metadata$treatment  # Replace 'treatment' with your grouping column


# Extract variance contributions
variance_explained <- spca_res$prop_expl_var$X * 100 # Convert to percentage

spca_scores_30$label <- ifelse(spca_scores_30$PC2 > 7, metadata$treatment, NA)
spca_scores_30$label2 <- ifelse(spca_scores_30$PC3 > 5, metadata$treatment, NA)
spca_scores_30$label3 <- ifelse(spca_scores_30$PC5 > 5 | spca_scores_30$PC5 < -4, metadata$treatment, NA)
spca_scores_30$label4 <- ifelse(spca_scores_30$PC7 > 5 | spca_scores_30$PC7 < -4, metadata$treatment, NA)
spca_scores_30$label5 <- ifelse(spca_scores_30$PC9 < -4 | spca_scores_30$PC10 < -3 | spca_scores_30$PC10 > 3, metadata$treatment, NA)

spca_scores_30$label7 <- ifelse(spca_scores_30$PC13 < -4 | spca_scores_30$PC13 > 5 | spca_scores_30$PC14 > 5 | spca_scores_30$PC14 < -5, metadata$treatment, NA)
spca_scores_30$label8 <- ifelse(spca_scores_30$PC15 > 4, metadata$treatment, NA)

# Plot PC1 vs PC2
ggplot(spca_scores_30, aes(x = PC1, y = PC2, color = group)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(
    title = "Sparse PCA: PC1 vs PC2",
    x = paste0("PC1 (", round(variance_explained[1], 2), "%)"),
    y = paste0("PC2 (", round(variance_explained[2], 2), "%)")
  ) + geom_text_repel(aes(label = label), show.legend = FALSE, max.overlaps = Inf)

# Plot PC3 vs PC4
ggplot(spca_scores_30, aes(x = PC3, y = PC4, color = group)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(
    title = "Sparse PCA: PC3 vs PC4",
    x = paste0("PC3 (", round(variance_explained[3], 2), "%)"),
    y = paste0("PC4 (", round(variance_explained[4], 2), "%)")
  ) + geom_text_repel(aes(label = label2), show.legend = FALSE, max.overlaps = Inf) 

# Plot PC5 vs PC6
ggplot(spca_scores_30, aes(x = PC5, y = PC6, color = group)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(
    title = "Sparse PCA: PC5 vs PC6",
    x = paste0("PC5 (", round(variance_explained[5], 2), "%)"),
    y = paste0("PC6 (", round(variance_explained[6], 2), "%)")
  ) + geom_text_repel(aes(label = label3), show.legend = FALSE, max.overlaps = Inf) 

# Plot PC7 vs PC8
ggplot(spca_scores_30, aes(x = PC7, y = PC8, color = group)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(
    title = "Sparse PCA: PC7 vs PC8",
    x = paste0("PC7 (", round(variance_explained[7], 2), "%)"),
    y = paste0("PC8 (", round(variance_explained[8], 2), "%)")
  ) + geom_text_repel(aes(label = label4), show.legend = FALSE, max.overlaps = Inf) 

# Plot PC9 vs PC10
ggplot(spca_scores_30, aes(x = PC9, y = PC10, color = group)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(
    title = "Sparse PCA: PC9 vs PC10",
    x = paste0("PC9 (", round(variance_explained[9], 2), "%)"),
    y = paste0("PC10 (", round(variance_explained[10], 2), "%)")
  ) + geom_text_repel(aes(label = label5), show.legend = FALSE, max.overlaps = Inf)

# Plot PC11 vs PC12
ggplot(spca_scores_30, aes(x = PC11, y = PC12, color = group)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(
    title = "Sparse PCA: PC11 vs PC12",
    x = paste0("PC11 (", round(variance_explained[11], 2), "%)"),
    y = paste0("PC12 (", round(variance_explained[12], 2), "%)")
  ) #+ geom_text_repel(aes(label = label6), show.legend = FALSE, max.overlaps = Inf) 

# Plot PC13 vs PC14
ggplot(spca_scores_30, aes(x = PC13, y = PC14, color = group)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(
    title = "Sparse PCA: PC13 vs PC14",
    x = paste0("PC13 (", round(variance_explained[13], 2), "%)"),
    y = paste0("PC14 (", round(variance_explained[14], 2), "%)")
  ) + geom_text_repel(aes(label = label7), show.legend = FALSE, max.overlaps = Inf) 

# Plot PC15 vs PC16
ggplot(spca_scores_30, aes(x = PC15, y = PC16, color = group)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(
    title = "Sparse PCA: PC13 vs PC14",
    x = paste0("PC15 (", round(variance_explained[15], 2), "%)"),
    y = paste0("PC16 (", round(variance_explained[16], 2), "%)")
  ) + geom_text_repel(aes(label = label8), show.legend = FALSE, max.overlaps = Inf) 

```

#### 30 genes per component
=======
```{r}
library(mixOmics)
library(GGally)
```

#### Calculate sparse PCA

Here, we will vary the number of genes per component to see what the results look like

```{r}
# Function to run sparse PCA and return scores and variance explained
run_spca <- function(hvgs, n_pcs, n_genes_per_component) {
  # Run sparse PCA
  spca_res <- spca(hvgs, ncomp = n_pcs, keepX = rep(n_genes_per_component, n_pcs))
  
  # Create a tibble with sample and PCs
  spca_scores <- as_tibble(spca_res$variates$X, rownames = "sample") %>%
    separate(sample, into = c("ligand", "group", "plate", "well"), sep = "_")
  
  # Extract variance contributions
  variance_explained <- spca_res$prop_expl_var$X * 100 # Convert to percentage
  
  # Return both outputs
  return(list(spca_scores_less = spca_scores, variance_explained = variance_explained))
}
```
>>>>>>> Stashed changes

```{r, fig.height = 12, fig.width=20}
if (ncol(hvgs) == 330) {
  hvgs <- t(hvgs)
}

<<<<<<< Updated upstream
# Run sparse PCA
spca_res_30 <- spca(hvgs, ncomp = 16, keepX = c(30, 30, 30, 30, 30, 30, 30, 30, 30, 30, 30, 30, 30, 30, 30, 30))  # Select 30 genes per component


#plotIndiv(spca_res, group = metadata$treatment, legend = TRUE)

spca_scores_30 <- as.data.frame(spca_res_30$variates$X)   # Samples x Components

# Add metadata to sPCA scores
spca_scores_30$group <- metadata$treatment  # Replace 'treatment' with your grouping column


# Extract variance contributions
variance_explained <- spca_res_30$prop_expl_var$X * 100 # Convert to percentage

spca_scores_30$label <- ifelse(spca_scores_30$PC2 > 5, metadata$treatment, NA)
spca_scores_30$label2 <- ifelse(spca_scores_30$PC3 > 5, metadata$treatment, NA)
spca_scores_30$label3 <- ifelse(spca_scores_30$PC5 > 4 | spca_scores_30$PC5 < -4, metadata$treatment, NA)
spca_scores_30$label4 <- ifelse(spca_scores_30$PC7 > 3 | spca_scores_30$PC7 < -4, metadata$treatment, NA)
spca_scores_30$label5 <- ifelse(spca_scores_30$PC9 < -4 | spca_scores_30$PC10 < -3 | spca_scores_30$PC10 > 3, metadata$treatment, NA)
spca_scores_30$label6 <- ifelse(spca_scores_30$PC12 < -4 | spca_scores_30$PC12 > 5 | spca_scores_30$PC11 < -4, metadata$treatment, NA)
spca_scores_30$label7 <- ifelse(spca_scores_30$PC13 < -4 | spca_scores_30$PC13 > 5 | spca_scores_30$PC14 > 5 | spca_scores_30$PC14 < -5, metadata$treatment, NA)
spca_scores_30$label8 <- ifelse(spca_scores_30$PC15 > 4, metadata$treatment, NA)

# Plot PC1 vs PC2
ggplot(spca_scores_30, aes(x = PC1, y = PC2, color = group)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(
    title = "Sparse PCA: PC1 vs PC2",
    x = paste0("PC1 (", round(variance_explained[1], 2), "%)"),
    y = paste0("PC2 (", round(variance_explained[2], 2), "%)")
  ) + geom_text_repel(aes(label = label), show.legend = FALSE, max.overlaps = Inf)

# Plot PC3 vs PC4
ggplot(spca_scores_30, aes(x = PC3, y = PC4, color = group)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(
    title = "Sparse PCA: PC3 vs PC4",
    x = paste0("PC3 (", round(variance_explained[3], 2), "%)"),
    y = paste0("PC4 (", round(variance_explained[4], 2), "%)")
  ) + geom_text_repel(aes(label = label2), show.legend = FALSE, max.overlaps = Inf) 

# Plot PC5 vs PC6
ggplot(spca_scores_30, aes(x = PC5, y = PC6, color = group)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(
    title = "Sparse PCA: PC5 vs PC6",
    x = paste0("PC5 (", round(variance_explained[5], 2), "%)"),
    y = paste0("PC6 (", round(variance_explained[6], 2), "%)")
  ) + geom_text_repel(aes(label = label3), show.legend = FALSE, max.overlaps = Inf) 

# Plot PC7 vs PC8
ggplot(spca_scores_30, aes(x = PC7, y = PC8, color = group)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(
    title = "Sparse PCA: PC7 vs PC8",
    x = paste0("PC7 (", round(variance_explained[7], 2), "%)"),
    y = paste0("PC8 (", round(variance_explained[8], 2), "%)")
  ) + geom_text_repel(aes(label = label4), show.legend = FALSE, max.overlaps = Inf) 

# Plot PC9 vs PC10
ggplot(spca_scores_30, aes(x = PC9, y = PC10, color = group)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(
    title = "Sparse PCA: PC9 vs PC10",
    x = paste0("PC9 (", round(variance_explained[9], 2), "%)"),
    y = paste0("PC10 (", round(variance_explained[10], 2), "%)")
  ) + geom_text_repel(aes(label = label5), show.legend = FALSE, max.overlaps = Inf)

# Plot PC11 vs PC12
ggplot(spca_scores_30, aes(x = PC11, y = PC12, color = group)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(
    title = "Sparse PCA: PC11 vs PC12",
    x = paste0("PC11 (", round(variance_explained[11], 2), "%)"),
    y = paste0("PC12 (", round(variance_explained[12], 2), "%)")
  ) + geom_text_repel(aes(label = label6), show.legend = FALSE, max.overlaps = Inf) 

# Plot PC13 vs PC14
ggplot(spca_scores_30, aes(x = PC13, y = PC14, color = group)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(
    title = "Sparse PCA: PC13 vs PC14",
    x = paste0("PC13 (", round(variance_explained[13], 2), "%)"),
    y = paste0("PC14 (", round(variance_explained[14], 2), "%)")
  ) + geom_text_repel(aes(label = label7), show.legend = FALSE, max.overlaps = Inf) 

# Plot PC15 vs PC16
ggplot(spca_scores_30, aes(x = PC15, y = PC16, color = group)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(
    title = "Sparse PCA: PC13 vs PC14",
    x = paste0("PC15 (", round(variance_explained[15], 2), "%)"),
    y = paste0("PC16 (", round(variance_explained[16], 2), "%)")
  ) + geom_text_repel(aes(label = label8), show.legend = FALSE, max.overlaps = Inf) 
=======
# Calculate sparse pca results over different gene numbers
spca_res_50 <- run_spca(hvgs, 24, 50)
spca_res_25 <- run_spca(hvgs, 24, 25)
spca_res_10 <- run_spca(hvgs, 24, 10)
spca_res_05 <- run_spca(hvgs, 24, 5)

# Access the outputs
spca_scores_50 <- spca_res_50$spca_scores
spca_scores_25 <- spca_res_25$spca_scores
spca_scores_10 <- spca_res_10$spca_scores
spca_scores_05 <- spca_res_05$spca_scores
variance_explained_50 <- spca_res_50$variance_explained
variance_explained_25 <- spca_res_25$variance_explained
variance_explained_10 <- spca_res_10$variance_explained
variance_explained_05 <- spca_res_05$variance_explained
>>>>>>> Stashed changes
```

#### Plot sparse PCA results

```{r}
# returns ggplot object
plot_spca <- function(
  spca_scores,  # Data frame containing PCA scores and associated metadata
  var_df,       # Data frame or named vector with variance explained for each PC
  pc_x,         # Name of the PC to plot on the x-axis (e.g., "PC1")
  pc_y,         # Name of the PC to plot on the y-axis (e.g., "PC2")
  pc_x_cutoffs = NA,  # Optional: Numeric vector with lower and upper cutoffs for pc_x for labeling
  pc_y_cutoffs = NA,  # Optional: Numeric vector with lower and upper cutoffs for pc_y  for labeling
  color_by = "group"  # Name of the column in spca_scores to use for coloring points
) {
  # Create a label column
  spca_scores <- spca_scores %>%
    mutate(label = ifelse((.data[[pc_x]] < pc_x_cutoffs[1]) |
                            (.data[[pc_x]] > pc_x_cutoffs[2]) |
                            (.data[[pc_y]] < pc_y_cutoffs[1]) |
                            (.data[[pc_y]] > pc_y_cutoffs[2]),
                          ligand, NA
    ))
  
  p <- ggplot(spca_scores, aes(x = .data[[pc_x]], y = .data[[pc_y]], color = .data[[color_by]])) +
    geom_point(size = 3, alpha = 0.7) +
    geom_text_repel(
      aes(label = label),
      show.legend = FALSE,
      max.overlaps = Inf,
      size = 2
    ) +
    labs(x = paste0(pc_x, " (", round(var_df[pc_x],2), "%)"),
         y = paste0(pc_y, " (", round(var_df[pc_y],2), "%)")) +
    theme(legend.position = "none", aspect.ratio = 1)
  
  return(p)
}
```

```{r}
# Create the list of PC pairs to plot
pcs <- paste0("PC", 1:n_pcs)
pc_pairs <- split(pcs, ceiling(seq_along(pcs)/2))

# create plot lists for each spca run
pcPlots50 <- vector(mode = "list", length = length(pc_pairs))
for(i in 1:length(pc_pairs)){
  pcPlots50[[i]] <- plot_spca(spca_scores_25,
                            variance_explained_25,
                            pc_x = pc_pairs[[i]][1],
                            pc_y = pc_pairs[[i]][2],
                            pc_x_cutoffs = c(-5,5),
                            pc_y_cutoffs = c(-5,5),
                            color_by = "plate")
}

pcPlots25 <- vector(mode = "list", length = length(pc_pairs))
for(i in 1:length(pc_pairs)){
  pcPlots25[[i]] <- plot_spca(spca_scores_25,
                            variance_explained_25,
                            pc_x = pc_pairs[[i]][1],
                            pc_y = pc_pairs[[i]][2],
                            pc_x_cutoffs = c(-5,5),
                            pc_y_cutoffs = c(-5,5),
                            color_by = "ligand")
}

pcPlots10 <- vector(mode = "list", length = length(pc_pairs))
for(i in 1:length(pc_pairs)){
  pcPlots10[[i]] <- plot_spca(spca_scores_10,
                            variance_explained_10,
                            pc_x = pc_pairs[[i]][1],
                            pc_y = pc_pairs[[i]][2],
                            pc_x_cutoffs = c(-5,5),
                            pc_y_cutoffs = c(-5,5),
                            color_by = "ligand")
}

pcPlots05 <- vector(mode = "list", length = length(pc_pairs))
for(i in 1:length(pc_pairs)){
  pcPlots05[[i]] <- plot_spca(spca_scores_05,
                            variance_explained_05,
                            pc_x = pc_pairs[[i]][1],
                            pc_y = pc_pairs[[i]][2],
                            pc_x_cutoffs = c(-5,5),
                            pc_y_cutoffs = c(-5,5),
                            color_by = "ligand")
}
```

```{r, fig.height=15, fig.width=20}
# plot 50 gene components
plot_grid(plotlist = pcPlots50)

# plot 25 gene components
plot_grid(plotlist = pcPlots25)

# plot 10 gene components
plot_grid(plotlist = pcPlots10)

# plot 5 gene components
plot_grid(plotlist = pcPlots05)
```



#### 10 genes per component

```{r, fig.height = 12, fig.width=20}
if (ncol(hvgs) == 310) {
  hvgs <- t(hvgs)
}

# Run sparse PCA
spca_res_10 <- spca(hvgs, ncomp = 16, keepX = c(10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10))  # Select 10 genes per component


#plotIndiv(spca_res, group = metadata$treatment, legend = TRUE)

spca_scores_10 <- as.data.frame(spca_res_10$variates$X)   # Samples x Components

# Add metadata to sPCA scores
spca_scores_10$group <- metadata$treatment  # Replace 'treatment' with your grouping column


# Extract variance contributions
variance_explained <- spca_res_10$prop_expl_var$X * 100 # Convert to percentage

spca_scores_10$label <- ifelse(spca_scores_10$PC2 > 5, metadata$treatment, NA)
spca_scores_10$label2 <- ifelse(spca_scores_10$PC3 > 5, metadata$treatment, NA)
spca_scores_10$label3 <- ifelse(spca_scores_10$PC5 > 4 | spca_scores_10$PC5 < -4, metadata$treatment, NA)
spca_scores_10$label4 <- ifelse(spca_scores_10$PC7 > 3 | spca_scores_10$PC7 < -4, metadata$treatment, NA)
spca_scores_10$label5 <- ifelse(spca_scores_10$PC9 < -4 | spca_scores_10$PC10 < -3 | spca_scores_10$PC10 > 3, metadata$treatment, NA)
spca_scores_10$label6 <- ifelse(spca_scores_10$PC12 < -4 | spca_scores_10$PC12 > 5 | spca_scores_10$PC11 < -4, metadata$treatment, NA)
spca_scores_10$label7 <- ifelse(spca_scores_10$PC13 < -4 | spca_scores_10$PC13 > 5 | spca_scores_10$PC14 > 5 | spca_scores_10$PC14 < -5, metadata$treatment, NA)
spca_scores_10$label8 <- ifelse(spca_scores_10$PC15 > 4, metadata$treatment, NA)

# Plot PC1 vs PC2
ggplot(spca_scores_10, aes(x = PC1, y = PC2, color = group)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(
    title = "Sparse PCA: PC1 vs PC2",
    x = paste0("PC1 (", round(variance_explained[1], 2), "%)"),
    y = paste0("PC2 (", round(variance_explained[2], 2), "%)")
  ) + geom_text_repel(aes(label = label), show.legend = FALSE, max.overlaps = Inf)

# Plot PC3 vs PC4
ggplot(spca_scores_10, aes(x = PC3, y = PC4, color = group)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(
    title = "Sparse PCA: PC3 vs PC4",
    x = paste0("PC3 (", round(variance_explained[3], 2), "%)"),
    y = paste0("PC4 (", round(variance_explained[4], 2), "%)")
  ) + geom_text_repel(aes(label = label2), show.legend = FALSE, max.overlaps = Inf) 

# Plot PC5 vs PC6
ggplot(spca_scores_10, aes(x = PC5, y = PC6, color = group)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(
    title = "Sparse PCA: PC5 vs PC6",
    x = paste0("PC5 (", round(variance_explained[5], 2), "%)"),
    y = paste0("PC6 (", round(variance_explained[6], 2), "%)")
  ) + geom_text_repel(aes(label = label3), show.legend = FALSE, max.overlaps = Inf) 

# Plot PC7 vs PC8
ggplot(spca_scores_10, aes(x = PC7, y = PC8, color = group)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(
    title = "Sparse PCA: PC7 vs PC8",
    x = paste0("PC7 (", round(variance_explained[7], 2), "%)"),
    y = paste0("PC8 (", round(variance_explained[8], 2), "%)")
  ) + geom_text_repel(aes(label = label4), show.legend = FALSE, max.overlaps = Inf) 

# Plot PC9 vs PC10
ggplot(spca_scores_10, aes(x = PC9, y = PC10, color = group)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(
    title = "Sparse PCA: PC9 vs PC10",
    x = paste0("PC9 (", round(variance_explained[9], 2), "%)"),
    y = paste0("PC10 (", round(variance_explained[10], 2), "%)")
  ) + geom_text_repel(aes(label = label5), show.legend = FALSE, max.overlaps = Inf)

# Plot PC11 vs PC12
ggplot(spca_scores_10, aes(x = PC11, y = PC12, color = group)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(
    title = "Sparse PCA: PC11 vs PC12",
    x = paste0("PC11 (", round(variance_explained[11], 2), "%)"),
    y = paste0("PC12 (", round(variance_explained[12], 2), "%)")
  ) + geom_text_repel(aes(label = label6), show.legend = FALSE, max.overlaps = Inf) 

# Plot PC13 vs PC14
ggplot(spca_scores_10, aes(x = PC13, y = PC14, color = group)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(
    title = "Sparse PCA: PC13 vs PC14",
    x = paste0("PC13 (", round(variance_explained[13], 2), "%)"),
    y = paste0("PC14 (", round(variance_explained[14], 2), "%)")
  ) + geom_text_repel(aes(label = label7), show.legend = FALSE, max.overlaps = Inf) 

# Plot PC15 vs PC16
ggplot(spca_scores_10, aes(x = PC15, y = PC16, color = group)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(
    title = "Sparse PCA: PC13 vs PC14",
    x = paste0("PC15 (", round(variance_explained[15], 2), "%)"),
    y = paste0("PC16 (", round(variance_explained[16], 2), "%)")
  ) + geom_text_repel(aes(label = label8), show.legend = FALSE, max.overlaps = Inf) 
```

#### 5 genes per component

```{r, fig.height = 12, fig.width=20}
if (ncol(hvgs) == 330) {
  hvgs <- t(hvgs)
}

# Run sparse PCA
spca_res_5 <- spca(hvgs, ncomp = 16, keepX = c(5, 5, 5, 5, 5, 5, 5, 5, 5, 5, 5, 5, 5, 5, 5, 5))  # Select 5 genes per component


#plotIndiv(spca_res, group = metadata$treatment, legend = TRUE)

spca_scores_5 <- as.data.frame(spca_res_5$variates$X)   # Samples x Components

# Add metadata to sPCA scores
spca_scores_5$group <- metadata$treatment  # Replace 'treatment' with your grouping column


# Extract variance contributions
variance_explained <- spca_res_5$prop_expl_var$X * 100 # Convert to percentage

spca_scores_5$label <- ifelse(spca_scores_5$PC2 > 5, metadata$treatment, NA)
spca_scores_5$label2 <- ifelse(spca_scores_5$PC3 > 5, metadata$treatment, NA)
spca_scores_5$label3 <- ifelse(spca_scores_5$PC5 > 4 | spca_scores_5$PC5 < -4, metadata$treatment, NA)
spca_scores_5$label4 <- ifelse(spca_scores_5$PC7 > 3 | spca_scores_5$PC7 < -4, metadata$treatment, NA)
spca_scores_5$label5 <- ifelse(spca_scores_5$PC9 < -4 | spca_scores_5$PC10 < -3 | spca_scores_5$PC10 > 3, metadata$treatment, NA)
spca_scores_5$label6 <- ifelse(spca_scores_5$PC12 < -4 | spca_scores_5$PC12 > 5 | spca_scores_5$PC11 < -4, metadata$treatment, NA)
spca_scores_5$label7 <- ifelse(spca_scores_5$PC13 < -4 | spca_scores_5$PC13 > 5 | spca_scores_5$PC14 > 5 | spca_scores_5$PC14 < -5, metadata$treatment, NA)
spca_scores_5$label8 <- ifelse(spca_scores_5$PC15 > 4, metadata$treatment, NA)

# Plot PC1 vs PC2
ggplot(spca_scores_5, aes(x = PC1, y = PC2, color = group)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(
    title = "Sparse PCA: PC1 vs PC2",
    x = paste0("PC1 (", round(variance_explained[1], 2), "%)"),
    y = paste0("PC2 (", round(variance_explained[2], 2), "%)")
  ) + geom_text_repel(aes(label = label), show.legend = FALSE, max.overlaps = Inf)

# Plot PC3 vs PC4
ggplot(spca_scores_5, aes(x = PC3, y = PC4, color = group)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(
    title = "Sparse PCA: PC3 vs PC4",
    x = paste0("PC3 (", round(variance_explained[3], 2), "%)"),
    y = paste0("PC4 (", round(variance_explained[4], 2), "%)")
  ) + geom_text_repel(aes(label = label2), show.legend = FALSE, max.overlaps = Inf) 

# Plot PC5 vs PC6
ggplot(spca_scores_5, aes(x = PC5, y = PC6, color = group)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(
    title = "Sparse PCA: PC5 vs PC6",
    x = paste0("PC5 (", round(variance_explained[5], 2), "%)"),
    y = paste0("PC6 (", round(variance_explained[6], 2), "%)")
  ) + geom_text_repel(aes(label = label3), show.legend = FALSE, max.overlaps = Inf) 

# Plot PC7 vs PC8
ggplot(spca_scores_5, aes(x = PC7, y = PC8, color = group)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(
    title = "Sparse PCA: PC7 vs PC8",
    x = paste0("PC7 (", round(variance_explained[7], 2), "%)"),
    y = paste0("PC8 (", round(variance_explained[8], 2), "%)")
  ) + geom_text_repel(aes(label = label4), show.legend = FALSE, max.overlaps = Inf) 

# Plot PC9 vs PC10
ggplot(spca_scores_5, aes(x = PC9, y = PC10, color = group)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(
    title = "Sparse PCA: PC9 vs PC10",
    x = paste0("PC9 (", round(variance_explained[9], 2), "%)"),
    y = paste0("PC10 (", round(variance_explained[10], 2), "%)")
  ) + geom_text_repel(aes(label = label5), show.legend = FALSE, max.overlaps = Inf)

# Plot PC11 vs PC12
ggplot(spca_scores_5, aes(x = PC11, y = PC12, color = group)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(
    title = "Sparse PCA: PC11 vs PC12",
    x = paste0("PC11 (", round(variance_explained[11], 2), "%)"),
    y = paste0("PC12 (", round(variance_explained[12], 2), "%)")
  ) + geom_text_repel(aes(label = label6), show.legend = FALSE, max.overlaps = Inf) 

# Plot PC13 vs PC14
ggplot(spca_scores_5, aes(x = PC13, y = PC14, color = group)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(
    title = "Sparse PCA: PC13 vs PC14",
    x = paste0("PC13 (", round(variance_explained[13], 2), "%)"),
    y = paste0("PC14 (", round(variance_explained[14], 2), "%)")
  ) + geom_text_repel(aes(label = label7), show.legend = FALSE, max.overlaps = Inf) 

# Plot PC15 vs PC16
ggplot(spca_scores_5, aes(x = PC15, y = PC16, color = group)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(
    title = "Sparse PCA: PC13 vs PC14",
    x = paste0("PC15 (", round(variance_explained[15], 2), "%)"),
    y = paste0("PC16 (", round(variance_explained[16], 2), "%)")
  ) + geom_text_repel(aes(label = label8), show.legend = FALSE, max.overlaps = Inf) 
```

# Fit LM for each PC

#make csv with all of the loadings for each pc

# for each pc, make boxplot of pc component

# try with different numbers of genes per component


```{r, fig.height=8, fig.width = 14}
# Helper function to generate boxplot for a specific PC
# pc index is character with PC name
generate_boxplot <- function(pc_index, spca_scores) {
  # Extract the PC scores for the specified PC
  pc_scores <- spca_scores[[pc_index]]
  boxplot_data <- data.frame(PC = pc_scores, Treatment = spca_scores$ligand)
  # Calculate the median PC score for each treatment
  median_pc <- tapply(pc_scores, boxplot_data$Treatment, median)
  # Order the treatment factor by the median PC score in descending order
  boxplot_data$Treatment <- factor(boxplot_data$Treatment,
                                   levels = names(sort(median_pc, decreasing = TRUE)))
  # Generate the boxplot using ggplot2
  p <- ggplot(boxplot_data, aes(x = Treatment, y = PC, fill = Treatment)) +
    geom_boxplot() +
    labs(
      x = "ligand",
      y = paste(pc_index, "Score")
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    theme(legend.position = "none")
  
  return(p)
}
```

```{r}
pcs <- paste0("PC", 1:20)
# Loop through PC1 to PC20 and store each plot in the list
pc_boxplot_50 <- list()
for (pc_index in pcs) {
  # Call the helper function for each PC and store the plot
  pc_boxplot_50[[pc_index]] <- generate_boxplot(pc_index, spca_scores_50)
}

pc_boxplot_25 <- list()
for (pc_index in pcs) {
  # Call the helper function for each PC and store the plot
  pc_boxplot_25[[pc_index]] <- generate_boxplot(pc_index, spca_scores_25)
}

pc_boxplot_10 <- list()
for (pc_index in pcs) {
  # Call the helper function for each PC and store the plot
  pc_boxplot_10[[pc_index]] <- generate_boxplot(pc_index, spca_scores_10)
}

pc_boxplot_05 <- list()
for (pc_index in pcs) {
  # Call the helper function for each PC and store the plot
  pc_boxplot_05[[pc_index]] <- generate_boxplot(pc_index, spca_scores_05)
}
```

```{r, fig.height=30, fig.width=40}
# plot 50 gene components
plot_grid(plotlist = pc_boxplot_50, ncol = 2)

# plot 25 gene components
plot_grid(plotlist = pc_boxplot_25, ncol = 2)

# plot 10 gene components
plot_grid(plotlist = pc_boxplot_10, ncol = 2)

# plot 5 gene components
plot_grid(plotlist = pc_boxplot_05, ncol = 2)
```

```{r}
# Helper function to generate dot plot for a specific PC
generate_dotplot <- function(pc_index, spca_res, metadata) {
  # Extract the PC scores for the specified PC
  pc_scores <- spca_res$x[, pc_index]
  
  # Create a data frame with PC scores and treatment information
  dotplot_data <- data.frame(PC = pc_scores, Treatment = metadata$treatment)
  
  # Calculate the median PC score for each treatment
  median_pc <- tapply(pc_scores, metadata$treatment, median)
  
  # Order the treatment factor by the median PC score in descending order
  dotplot_data$Treatment <- factor(dotplot_data$Treatment, levels = names(sort(median_pc, decreasing = TRUE)))
  
  # Generate the dot plot using ggplot2 with jitter for better visualization
  p <- ggplot(dotplot_data, aes(x = Treatment, y = PC, color = Treatment)) +
    geom_jitter(width = 0.2, size = 3, alpha = 0.7) +  # Jitter for dot visualization
    labs(
      title = paste("Dot Plot of PC", pc_index, "Scores Grouped by Treatment (Sorted)"),
      x = "Treatment",
      y = paste("PC", pc_index, "Score")
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
  
  return(p)
}

# List to store the plots
pc_dot_plots <- list()

# Loop through PC1 to PC16 and store each dot plot in the list
for (pc_index in 1:16) {
  # Call the helper function for each PC and store the plot
  pc_dot_plots[[pc_index]] <- generate_dotplot(pc_index, spca_res, metadata)
}

# Optionally: Print the first plot to check
print(pc_dot_plots[[2]])

# To access any plot, use pc_dot_plots[[pc_index]], for example:
# print(pc_dot_plots[[2]]) for PC2 plot
```

### Extract PC loadings and save to csv

```{r}
# Helper function to save loadings of all PCs to a CSV file
save_spca_loadings_to_csv <- function(spca_res, output_file = "spca_loadings.csv") {
  # Extract the loadings matrix from the SPCA result
  loadings_matrix <- spca_res$rotation
  
  loadings_matrix <- loadings_matrix[rowSums(loadings_matrix != 0) > 0, ]
  
  # Convert the loadings matrix into a data frame for easy manipulation
  loadings_df <- as.data.frame(loadings_matrix)
  
  # Add gene names (row names) as a column in the data frame
  loadings_df$ensembl_ID <- rownames(loadings_matrix)
  
  loadings_df <- merge(loadings_df, featureNames, by = "ensembl_ID", all.x = TRUE)
  
  # Reorder the columns so that Ensembl IDs and gene names appear first
  loadings_df <- loadings_df[, c("ensembl_ID", "gene", colnames(loadings_matrix))]
  
  # Reorder the columns so that 'ensembl_ID', 'gene', and PCs are in the correct order
  pc_columns <- grep("^PC\\d+$", colnames(loadings_df), value = TRUE) # Find PC columns
  loadings_df <- loadings_df[, c("ensembl_ID", "gene", pc_columns)]
  
  # Write the loadings data frame to a CSV file
  write.csv(loadings_df, file = output_file, row.names = FALSE)
  
  cat("Loadings CSV saved as:", output_file, "\n")
}

# Call the helper function to save loadings to CSV

# save_spca_loadings_to_csv(spca_res, "../analysis_outs/spca_loadings_50genes.csv")
# save_spca_loadings_to_csv(spca_res = spca_res30, "../analysis_outs/spca_loadings_30genes.csv")
# save_spca_loadings_to_csv(spca_res = spca_res10, "../analysis_outs/spca_loadings_10genes.csv")
# save_spca_loadings_to_csv(spca_res = spca_res5, "../analysis_outs/spca_loadings_5genes.csv")
```

