---
title: "Correlation Heatmaps"
author: "Ian Zumpano"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    html_preview: false
  html_notebook:
    toc: true
    toc_float: true
---

```{r setup}
library(tidyverse)
library(ggplot2)
library(cowplot)
library(patchwork)
library(DESeq2)
library(ComplexHeatmap)
library(IHW)
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
source("../functions/plotting_fxns.R")
theme_set(theme_Publication())
```

### [Import Data]{.underline}

```{r}
data <- read.csv("../processing_outs/count_matrix_umiDeDup_SIG06.csv", row.names = "X")

metadata <- read_csv("../processing_outs/processed_metadata_SIG06.csv")

# import feature names
featureNames <- read_csv("../processing_outs/featureNames_SIG06.csv")
featureNames <- select(featureNames, -category)
```

### [Normalization and Diagnostic Plots]{.underline}

```{r}
# prepare metadata for DESeq
metaDF <- as.data.frame(metadata[,-1])
rownames(metaDF) <- metadata$sample_ID

# match matrix and metadata
data <- data[,rownames(metaDF)]
```

#### Read normalized counts

```{r}
normCounts <- read.csv("../analysis_outs/counts_norm_DEseq2.csv")
```

#### Create DESeq object

```{r}
# import dds object
dds <- readRDS("/Users/izumpano/Desktop/Eric Analyses/SIG06/analysis_outs/dds_object_SIG06.rds")
```

#### Create Treatment Groups 

```{r}
# select viral treatments groups
viralGroups <- dds$treatment %>% unique()
viralGroups <- viralGroups[grep("recomb|none|linker", viralGroups, invert = T)]
viralGroups
```

```{r}
# select recombinant treatments groups
recombGroups <- dds$treatment %>% unique()
recombGroups <- grep("recomb", recombGroups, value = T)
recombGroups
```

```{r}
# select linkers
linkerGroups <- dds$treatment %>% unique()
linkerGroups <- grep("linker", linkerGroups, value = T)
linkerGroups
```

### Load DEG Analysis

```{r}
res_viral <- read.csv("../analysis_outs/res_viral.csv")
```

### Plot Correlation Heatmaps

### Prepare annotation and metadata

```{r}
metadata$treatment_type <- ifelse(metadata$treatment == "linker", "Linker",
                                    ifelse(metadata$treatment == "none", "None", "Other")) %>%  as_factor()

metadata$log_nCount_RNA <- log10(metadata$nCount_RNA)

annotation <- metadata %>% 
	select(sample_ID, replicate, plate, treatment_type, log_nCount_RNA) %>% data.frame(row.names = "sample_ID")
  
normCounts_rows <-normCounts %>% column_to_rownames("ensembl_ID") %>% select(-gene)

  
```

#### Create Helper Function to Plot Heatmaps

```{r}
# Define annotation colors
annotation_colors <- list(
  replicate = c(group1 = "darkblue", group2 = "darkgreen", group3 = "darkred"),
  plate = c(plate1 = "purple", plate2 = "orange", plate3 = "yellow", plate4 = "black"),
  treatment_type = c(Linker = "blue", None = "red", Other = "gray")
)

dark2_pal <- brewer.pal(8, "Dark2")

plot_ligand_correlation_heatmap <- function(norm_counts, metadata, annotation, method = "pearson", color_palette = "RdYlBu", title = "") {
  # Compute correlation matrix
  correlation_matrix <- cor(norm_counts, method = method)
  
  # Plot heatmap
  pheatmap(
    correlation_matrix,
    annotation_col = annotation,
    annotation_row = annotation,
    annotation_colors = annotation_colors,
    show_rownames = FALSE,
    show_colnames = FALSE,
    color = colorRampPalette(brewer.pal(11, color_palette))(100),
    cluster_rows = TRUE,
    cluster_cols = TRUE,
    legend = TRUE,
    main = paste(method, "Correlation Heatmap of Ligands", title)
  )
}
#plot_ligand_correlation_heatmap(normCounts_rows, metadata)

#normCounts_zscored <-zscored_data_group %>%select(-gene)

#plot_ligand_correlation_heatmap(zscored_data_group)

plot_ligand_correlation_heatmap(normCounts_vst, metadata)


gene_means <- rowMeans(normCounts_vst)
gene_variances <- apply(normCounts_vst, 1, var)

# You can define HVGs based on variance-to-mean ratio or top quantiles
variance_to_mean_ratio <- gene_variances / gene_means

# Define a threshold for selecting HVGs
threshold <- quantile(variance_to_mean_ratio, 0.9)  # Top 25% most variable genes
highly_variable_genes <- names(variance_to_mean_ratio)[variance_to_mean_ratio > threshold]

# Subset your data to include only highly variable genes
normCounts_HVG <- normCounts_vst[highly_variable_genes, ]

dim(normCounts_HVG)

top_ligands <- res_viral_sig %>% filter(padj < 0.05, log2FoldChange > 0.5) %>% pull(treatment) %>% unique()

#just need to filter the rownames to be within the ligands list

normCounts_DEG <- normCounts_vst[grepl(paste(top_ligands, collapse = "|"), sub("_.*", "", rownames(normCounts_vst))), ]



plot_ligand_correlation_heatmap(normCounts_DEG, metadata)

metadata

#only plot ligands that have DEgenes
```

#### Heatmap for all ligands using normalized counts

```{r, fig.width=18, fig.height=12}
plot_ligand_correlation_heatmap(normCounts_rows, metadata, annotation = annotation,  title = "(Normalized Counts)")
```

#### Heatmap for all ligands using vst normalized counts

```{r, fig.width=18, fig.height=12}
dds_norm <- vst(dds)
normCounts_vst <- assay(dds_norm)

plot_ligand_correlation_heatmap(normCounts_vst, metadata, annotation = annotation, title = "(VST Normalized Counts)")
```

#### Heatmap of only ligands with DEGs

```{r, fig.width=18, fig.height=12}
library(stringi)

# Create vector of ligands corresponding to DEGs
top_ligands <- res_viral %>% filter(padj < 0.01) %>% pull(treatment) %>% unique()

# filter for columns within normCounts that contain a ligand in top_ligands
pattern <- paste(top_ligands, collapse = "|")
cols  <- stri_detect_regex(names(normCounts_rows), pattern)
normCounts_DEG <- normCounts_rows[, cols]
dim(normCounts_DEG)


annotation_DEG <- annotation[rownames(annotation) %in% colnames(normCounts_DEG), , drop = FALSE]

plot_ligand_correlation_heatmap(normCounts_DEG, metadata, annotation = annotation_DEG, title = "(Top DEG Ligands)")


```

